Collaborative Visual Analysis on the Web with RCloud

### 2015-06-15 introduction.tex
working

More than a half-century ago, Tukey foresaw
much of what is now commonplace in data analysis~\cite{TukeyFDA}.
Powerful, interactive environments for analysis and programming were
the goal, together with an unflinching (and, at the time, somewhat
heretical) insistence in keeping humans as a central part of the
discovery process. His now-famous quip that ``the picture-examining eye is the
best finder we have of the wholly unanticipated'' has come to define
much of visual analytics and exploratory visualization~\cite{TukeyEDA}.

In some way, we have moved far beyond what Tukey imagined:
computing and networking capabilities today far exceed
what was barely imaginable then. 
We argue, on the other hand, that how we \emph{develop} our data analysis
solutions has not changed as much: the S language
was developed essentially alongside Unix~\cite{TheSSystem} and,
although environments such as RStudio include a variety of modern
features, they still follow the basic metaphor of terminal, text
editor, and source files stored in file systems.
We turn our attention to this opportunity to use computation
to support, not only individuals, but entire teams and their
work within larger organizations. In this paper, we contribute the
design of RCloud, together with an interview study conducted over the
course of its development and deployment at AT\&T Labs, where RCloud has
been in use for about two years.

Consider the role of a data science team within
a business or technical organization today.
Project teams vary in size, from just a few people to dozens or
more, even within one project's duration. Assignments are often
broad, and include tasks such as problem identification,
data wrangling, modeling, analysis, visualization, summarization,
presentation and interpretation of results, and recommending
actions to help clients to realize the benefits of the analysis.
Eventually, knowledge or working prototypes created by these teams
are transferred to other organizations to employ them in production.
There are many details to these tasks. For example, data wrangling can involve
finding data, understanding its syntax and semantics, assessing data quality,
performing normalization and data quality remediation, and making the data
available to other tasks that will follow.

Visual analytics depends greatly on communication and collaboration.
At almost every step, detailed knowledge about data, code and tasks
is shared by collaborators. 
Further, data scientists are increasingly asked to work more closely
with business or domain specialists who may be less technically oriented.
Thus, data scientists and developers are being asked to become
very broad, and integrate work across the spectrum.

%% Interviews in {\it Data Scientists at Work} by Sebastian Gutierrez illuminate
%% the teamwork and sharing in the current practice of visual analytics in industry.
%% We particularly appreciated the remarks by Jonathan Lenaghan about the potential
%% for continuous integration and publication.

%% % \begin{addmargin}[1em]{1em}  % left and right margin change
%% % part1.txt
%% Erin Shellman [of Nordstrom] described the benefits of
%% continuous deployment of experiments, and sharing knowledge
%% through software and data analysis artifacts.
%% ``Finally, prototyping our products so that internal
%% customers can use them early on has been crucial for
%% our success. It doesn't even have to be something
%% fancy. For instance, our recommendations preview
%% tool doesn't have a particularly interesting
%% visualization, but its enough to show the result of our
%% algorithms. Now we can shoot off a URL to internal
%% customers and it allows them to sit at their desk and
%% understand the behavior of our product and
%% experiment with it, and provide feedback way before
%% we're talking about getting it into production. This
%% has been super helpful and has been a really great
%% way to get people excited about what we're working
%% on...
%% %Building and maintaining those relationships are
%% %just like anything else-- you always have to be
%% %working on them. Nurturing those relationships are
%% %part of the job, and if you leave them unattended,
%% %they might not be there later.
%% %
%% % part2.txt
%% %Gutierrez: How do you share the knowledge you are
%% %building with others?
%% We’re obsessed with Confluence from
%% Atlassian, and the Data Lab has a very active
%% Confluence space. Recommendo is fully documented
%% on Confluence, so anyone in the company could learn
%% how it’s built, where to find it, and how to use it.
%% We also share exploratory analyses and reports on
%% Confluence so that we can still exchange knowledge
%% even if the work didn't make it into a larger project.''

%% % part3.txt
%% John Foreman commented on the scope of the work in
%% data science teams.
%% ``I lead the data science team at
%% MailChimp, and I like to get my hands dirty too.
%% Some of the big pieces of my day involve working
%% with my team to take stock of current projects and
%% figure out where to go next, doing my own work and
%% prototyping things-- I do projects just like my peers
%% do and then also my talking with other teams,
%% talking with management, and planning for the
%% future.
%% On our team, we've got different folks facing different
%% % part4.txt
%% kinds of projects. We've got one person who really
%% owns compliance and looks at our compliance
%% processes. We've got another data scientist who
%% focuses on more of the user experience side of the
%% house and understanding our customers. I help them
%% and others as is needed while I do the three things
%% mentioned earlier-- build data products, be a
%% translator, and have conversations with the data
%% science community.''

%% % part5.txt
%% Jonathan Lenaghan from PlaceIQ discussed to the potential
%% for continuous deployment in data science.
%% ``If you look at dev ops right now, they have things
%% such as continuous integration, continuous build,
%% automated testing, and test harnesses-- a1l of which
%% map very well from the dev ops world to the data ops
%% (a phrase I stole from Red Monk) world very easily. I
%% think this is a very powerful notion. It is important to
%% have testing frameworks for all of your data, so that if
%% you make a code change, you can go back and test all
%% of your data.''

%% % part6.txt
%% Anna Smith from Bitly spoke about the benefits of access to
%% a rich environment of shared data and code.
%% ``In addition to Bitly's extraordinary data sets,
%% I was given all the resources that I needed. These were
%% resources that I’d only heard about but had never
%% seen. I went from asking what a Hadoop cluster was,
%% as my school didn't have one, to being able to work
%% with one. Now I could play with one to figure out how
%% to make my work better and how to change my
%% algorithms so they were cleaner and ran faster on
%% Hadoop. The whole transition was eye opening. Not
%% only that, I could also look at other people’s code and
%% play with their code and data sets as well. It was
%% fantastic and I was convinced that working with data was my thing.''
%% % \end{addmargin}

%% % (Hiring ``full stack developers'' to perform integration seems to be a trend, but does not achieve the potential benefits of specialization and division of labor in larger teams, and is not a practice to be encouraged.)

Unfortunately, the processes and technologies supporting data analysis visual analytics are fragmented.
Knowledge is shared through informal conversations, emails,
meetings, phone calls, source code, in wikis, project documents and
by other means. Results of experiments are often shared first by copying static
output, usually a screen shot or image output. By the time
decisions are made based on the results of that screenshot, data
and processes may have changed so much that the decision can be wrong.

%% This disjointed, inconsistent process inhibits the use of tools to bridge
%% the many gaps and to help people find resources.
%% For example, even when code is shared, we still not may know where to
%% find examples that show how to use it.
%% When we find a useful data source, we don't know what packages or
%% other data sources usually go along with it, what problems with the data
%% others have already discovered and corrected, or where to obtain online feeds
%% to get updates.

This situation can affect how data science teams work; as analysts and
tool builders, we have experienced this first-hand in our environment, and
others report similar findings. Gutierrez collected interviews with data
scientists~\cite{Gutierrez:2014:DSA}, which include the following:

@@

Currently, these concerns are being addressed \emph{alongside} the
visual analytics environment. Some symptoms of the situation are:
\begin{itemize}
\item It is hard to find data, metadata, and knowledge about them.
\item Most coordination is done through meetings, whose content is not linked to other artifacts and may not even be stored.
\item Production deployment requires porting code to a different environment or even completely rewriting it, thwarting continuous release.
\item Many tools adopt a standalone or single developer perspective,
not suited to collaboration and web deployment. 
\item Analysts find insufficient support for tracing events or issues from production back to the EDA process.
\end{itemize}

In view of the opportunity to improve this situation,
we created a software environment that supports the end-to-end
visual analytics process for individuals and teams.
This environment knits together some familiar tools, and 
provides new features to find data and code; create experimental workbooks;
run experiments; annotate and deploy experiments as end-user websites or
as reusable, callable services; and to share, search and recommend these artifacts. The artifacts are stored in a version
control system that provides a common workspace, as well as needed control
and isolation between stable and experimental versions of code and other
resources.

@@

From a more principled or theoretical perspective, certain visual analytics
goals or objectives for emerging systems, described in previous work,
closely match ours. We identified the following central themes.

\begin{enumerate}

\item Technology transfer.
In most organizations, development of analyses and visualizations 
s done by \emph{hackers} and \emph{scripters} (adopting Kandel et al.'s
terminology~\cite{Kandel:2012:EDA}). \emph{Application users} 
gain the benefits of the tools developed by hackers and scripters.
Generally, the connection between these two worlds is made by IT staff,
who port or even rewrite code so it can run as a stable production service.
In an environment where business needs can change rapidly, this
process does not scale. Our objective is to merge the worlds
of EDA and production.

\item Coexistence. In the current data science ecosystem, the
isolation of exploratory visualization and data analysis
environments hinders wider adoption of modern techniques from each.
The richness of interactive visualization tools is still somewhat
separated from the power of statistical programming environments.
There is an opportunity to provide a framework so that developments
on each side are more easily adopted by the other, and made available
in production services.

\item Discoverability. A chronic complaint of data analysis teams is
repeated work (``How can I connect to database X?'', ``How do you get
clean data from column Y from feed Z?'') This work arises from lack of
communication about previously solved subproblems. Our goal is encourage
and support transparency of work between team members.

@@

%\carlos{we need a segue into related work}.
%\stephen{maybe not?}

>>>>

More than a half-century ago, Tukey foresaw
much of what is now commonplace in data analysis~\cite{TukeyFDA}.
Powerful, interactive environments for analysis and programming were
the goal, with unswerving insistence on visualization as a
central part of the discovery process. His now-famous quip that
``the picture-examining eye is the best finder we have of the
wholly unanticipated'' has come to define much of visual analytics
and exploratory visualization~\cite{TukeyEDA}.

Computing has moved far beyond what Tukey imagined.
Processing and networking capabilities today far exceed
what was barely imaginable then. Sophisticated technologies
for processing, storing, publishing, searching and sharing
information and analyses are commonplace.
But how we \emph{develop} analytic solutions has not changed much. 
Although environments such as RStudio include many features to
graphically interact with a local workspace, still follow the metaphor
of terminal, text editor, and source files stored in file systems.

The work of data scientists in teams is also changing.
Project teams vary in size, from just a few people to dozens or
more.
Assignments are broad and include tasks such as problem identification,
data wrangling, modeling, analysis, visualization, summarization,
presentation and interpretation of results, and recommending
actions to help clients to realize the benefits of the analysis.
Knowledge or software prototypes created by these teams are transferred
to other organizations to employ them in production.
This work depends extensively on communication and collaboration.
At almost every step, collaborators share detailed knowledge about tasks,
data and code.
Further, data scientists are increasingly asked to work closely
with less technically-oriented business or domain specialists.

Previous work has identified important areas for improvement in
visual analytics systems to address this situation.

\begin{enumerate}
\item Discoverability. A chronic complaint of data analysis teams is
inefficiency due to a lack of transparency in accessing the work of
other team memnbers.  The processes and technologies supporting
data analysis and visual analytics today are fragmented.
Knowledge is shared imperfectly through informal conversations, emails,
meetings, phone calls, source code, wikis and project documents.
This causes work to be repeated unnecessarily
(``How can I connect to database X?'', ``How do you get clean data
from column Y from feed Z?'') due to lack of communication about
previously solved problems. 
Our goal is support and
promote transparency of work between team members.

Results of experiments are often shared first by copying screen
shots or static output. By the time decisions are made based on
the results of that screenshot, data and processes may have changed
so much that the decision can be wrong.

\item Technology transfer.
In most organizations, development of analyses and visualizations 
s done by \emph{hackers} and \emph{scripters} (adopting Kandel et al.'s
terminology~\cite{Kandel:2012:EDA}). \emph{Application users} 
gain the benefits of the tools developed by hackers and scripters.
Results of experiments are often shared first by copying screen
shots or static output.  By the time decisions are made based on
the results of that screenshot, data and processes may have changed
so much that the decision can be wrong.
\stephen{Transitional sentence here.}
Generally, the connection between these two worlds is made by IT staff,
who port or rewrite code so it can run as a stable production service.
In an environment where business needs can change rapidly, this
process does not scale. Our goal is to merge the worlds of EDA and
production.

\item Coexistence. In current ecosystems, the isolation of exploratory
visualization and data analysis environments hinders wider adoption of
modern techniques from each.
The richness of interactive visualization tools is still somewhat
separated from the power of statistical programming environments.
There is an opportunity to provide a framework so that developments
on each side are more easily adopted by the other, and made available
in production services.

Fortunately, there is abundant evidance that technology can
solve many of these problems.
Gutierrez collected interviews with data scientists~\cite{Gutierrez:2014:DSA},
which include the following:

@@

\item Many tools adopt a standalone or single developer perspective,
not suited to collaboration and web deployment. 
\item Analysts find insufficient support for tracing events or issues from production back to the EDA process.
\end{itemize}

In view of the opportunity to improve this situation,
we created a software environment that supports the end-to-end
visual analytics process for individuals and teams and
their work within larger organizations.

In this paper, we contribute the
design of RCloud, with an interview study based on the
course of its development and deployment at AT\&T Labs for 
about three years.

This environment knits together some familiar tools, and 
provides new features to find data and code; create experimental workbooks;
run experiments; annotate and deploy experiments as end-user websites or
as reusable, callable services; and to share, search and recommend these artifacts. The artifacts are stored in a version
control system that provides a common workspace, as well as needed control
and isolation between stable and experimental versions of code and other
resources.

@@

@@
